Supplementary Information

### 2020-09-10 doc/manuscript/MEE_manuscript.Rmd
found Sheil paper!
Zofin corrections

Kristina J. Anderson-Teixeira^1,2^*,
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Camille Piponiot^1,2^,
Ross Alexander,
Craig D. Allen,
Raquel Alfaro-Sánchez,
Tala Awada,
Jennifer Baltzar,
Patrick Baker,
Sarayudh Bunyavejchewin,
Paolo Cherubini,
Justin Cooper,
Stuart Davies^2^,
Ryan Helcoski^1^,
Vasickova Ivana,
Kaspar Jakub,
Kamil Kral,
James Lutz,
Ellis Q. Margolis,
Justin Maxwell,
Sean McMahon^2^,
Helene Muller-Landau^2^,
Samonil Pavil,
Sabrina Russo,
Anastasia Sniderhan,
Alan J. Tepley^1,#^,
Mart Vlam,
Peter Zuidema

@@

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. **2**, **S1**; Table S5; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*. $T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig. **3**).
Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites). 
Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
At two of the higher-latitude temperate sites (HF and Zofin), temperatures were most influential during late winter/ early spring.
There were also a few instances where previous growing season conditions had the strongest influence.

@@

**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. **4**).** 
While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. **4**).
On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig. **4**).
At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.

@@

**The diversity of growth trends in relation to $DBH$ observed here (Fig. **4**) is largely attributable to species ecology and stand history (Fig. 4).**
On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by exponential decline. 
The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions.
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (DENDRO_REFS).
On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Zofin) exhibited initially low, but increasing, $\Delta r$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $\Delta r$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]. 
While the low community mean $\Delta r$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $\Delta r$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
Thus, patterns of decreasing $\Delta r$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
--  (*cite paper that I reviewed several years back; [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).

@@

### Acknowledgements
Thanks to Pete Kerby-Miller for bark thickness data.
Helpful feedback was provided by Albert Kim...
This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.

>>>>

Kristina J. Anderson-Teixeira^1,2^*,
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Camille Piponiot^1,2^,
Ross Alexander,
Craig D. Allen,
Raquel Alfaro-Sánchez,
Tala Awada,
Jennifer Baltzar,
Patrick Baker,
Sarayudh Bunyavejchewin,
Paolo Cherubini,
Justin Cooper,
Stuart Davies^2^,
Ryan Helcoski^1^,
Jakub Kašpar^cz^,
Kamil Kral,
James Lutz,
Ellis Q. Margolis,
Justin Maxwell,
Sean McMahon^2^,
Helene Muller-Landau^2^,
Sabrina Russo,
Pavel Šamonil^cz^,
Anastasia Sniderhan,
Alan J. Tepley^1,#^,
Ivana Vašíčková^cz^, 
Mart Vlam,
Peter Zuidema

@@

cz. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. **2**, **S1**; Table S5; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*. $T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig. **3**).
Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites). 
Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
There were also a few instances where previous growing season conditions had the strongest influence.

@@

**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. **4**).** 
While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. **4**).
On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig. **4**).
At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.

@@

**The diversity of growth trends in relation to $DBH$ observed here (Fig. **4**) is largely attributable to species ecology and stand history (Fig. 4).**
On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by exponential decline. 
The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions.
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (DENDRO_REFS).
On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $\Delta r$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $\Delta r$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]. 
While the low community mean $\Delta r$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $\Delta r$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
Thus, patterns of decreasing $\Delta r$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
--  (*cite Sheil et al. 2017 [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).

@@

### Acknowledgements
Thanks to Pete Kerby-Miller for bark thickness data.
Helpful feedback was provided by Albert Kim...
This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S
